Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-31 Report/Sections/data.tex
Stylistic changes to report

To illustrate the process of constructing a model using HMM building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.

Understanding animal behaviour can be important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.

Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the true energetic requirements of NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range are ignored as well. A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data are smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contains a total of 267 dives. Figure \ref{fig:data} displays the dive profile and accelerometer data.

Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, $Y$ is a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. The fine-scale process $Y^*$ is the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.

@@

The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics. A window size of two-second window is selected since it appears be the smallest window size that captures the ``wiggles" of the acceleration data in their entirety. Because the data was collected at the rate of 50 Hz, each two-second window contains a total of 100 raw acceleration readings. The acceleration data was also tri-axial, or collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than Equation (\ref{eqn:z}):
%
\begin{equation*}
    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right), \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation*}
%
Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar. We calculated $Z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1). Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$. In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained because subsequent observations share a subdive state $X^*_{t,t^*}$. Therefore, a simple HMM is used to model $Z^{*(2)}$. Given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%

We do not use information criteria to select the number of dive types and subdive types since they tend to overestimate the number of states in biological processes \citep{Pohle:2017}. If the hidden states are well-separated, a lag plot will reveal $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state. This is unfortunately not the case for the killer whale data, but we still consult these lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors. The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM. We force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that dive types 1 and 2 share subdive states. The parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
Remember that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed so far. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT. Additionally, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM. However, the Fourier sums $Z^{*(2)}$ are omitted from this model altogether. The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have the same emission distributions as the CarHHMM-DFT when applicable. Each of the three models leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier sums $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.

>>>>

To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale and the distribution of acceleration within each behavioral state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range are ignored as well. A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data are smoothed by taking a moving average within a window of 1/10th of a second. Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, $Y$ is a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. The fine-scale process $Y^*$ is the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.

@@

The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics. A window size of two-seconds is selected since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data. The acceleration data was also collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than in Equation (\ref{eqn:z}):
%
\begin{equation*}
    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right), \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation*}
%
Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar. We calculate $Z_t^{*(2)}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1). Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$. In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$. In words, each axis of acceleration has its own mean and variance, but all axes share an auto-correlation parameter.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$. Therefore, a simple HMM is used to model $Z^{*(2)}$. Given that $X^*_{t,t^*} = i^*$, the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%

We do not use information criteria to select the number of dive types and subdive types since they tend to overestimate the number of states in biological processes \citep{Pohle:2017}. If the hidden states are well-separated, a lag plot reveals $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state \cite{Lawler:2019}. This is unfortunately not the case for our killer whale data, but we still consult our lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors. The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM. We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that dive types 1 and 2 share subdive states. In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
Recall that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed so far. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT. In addition to the CarHHMM-DFT we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM. However, the Fourier sums $Z^{*(2)}$ are omitted from this model altogether. The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have emission distributions identical to the CarHHMM-DFT when applicable. Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.

### 2020-07-31 Report/Sections/discussion.tex
Stylistic changes to report

Current FDA literature addresses sequences of curves either as multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009} or as functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}. However, many real-world curve sequences exhibit both temporal dependence and a discrete number of types. One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with running, walking, etc. \citep{Morris:2007}.

To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves. Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation within a sequence of curves. One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions. The parameters of this fine-scale model then in turn depend upon the curve type. Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously and at different time-scales.

@@

Several possible choices exist for the fine-scale model, including yet another HMM. \citet{Barajas:2017} introduce this particular situation as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe apparent periodic behavior in the acceleration data. This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.

@@

To deal with these problematic fine-scale processes, we suggest using a moving-window transformation over the fine-scale observations. For periodic behaviour in particular, we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale. On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (possibly using a lag plot) calls for it. After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT. This model effectively decodes both dive types and subdive states of marine mammals while also taking into account the connection between these two levels.

This solves an important problem, as the energetic requirements of killer whales together with their energetic intake from predation is directly related to their survival \citep{Williams:2009,Noren:2011}. Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}. Accelerometers are an appealing alternative, but require decoding the behavioral state of the animal in order to effectively estimate energy expenditure \citep{Dot:2016}. The CarHHMM-DFT solves this problem by decoding both the subdive behavioral state and the dive type of a killer whale. It also describes the distribution of acceleration in the states. All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.

While we combine these models together effectively for our case study, in general this practice should be done with care. There are drawbacks to building arbitrarily complex models with HMMs. For example, HMMs require observations to be taken at regular intervals. Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue, but we do not describe them in detail here. This is because most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit. In addition, we show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions. This implies that a CarHMM is interpretable like a continuous-time model, but as easy to fit as a traditional HMM.

Another consideration when using HMMs to build complex models is the importance of effectively capturing the process being modelled while avoiding over-fitting and slow parameter estimation. One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states (i.e. subdive state 1 is the same across all dive types). Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states. However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

@@

Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models. When model complexity is kept under control, the resulting hierarchical model is both highly flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale. It is our hope that researchers adapt the methodology described here to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.

>>>>

Current FDA literature addresses sequences of curves either with multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009}, or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}. However, many real-world curve sequences exhibit both temporal dependence and a discrete number of types. One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with running, walking, etc. \citep{Morris:2007}.

In this work, we deal with both temporal dependence between curves and a discrete number of curve types by modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves. Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation between curves. One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions. The parameters of this fine-scale model then in turn depend upon the curve type. Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.

@@

Several possible choices exist for the fine-scale model, including yet another HMM. \citet{Barajas:2017} introduce this particular situation as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. Traditional HMMs prove insufficient at these very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe periodic behavior in the acceleration data. This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.

@@

To deal with these problematic fine-scale processes, we detail a moving-window transformation over the fine-scale observations. For periodic behaviour in particular, we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale. On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it. After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT. This model effectively decodes both dive types and subdive states of a killer whale while also taking into account the interaction between these two levels.

This solves an important problem, as the energetic requirements of killer whales together with their energetic intake from predation is directly related to their survival \citep{Williams:2009,Noren:2011}. Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}. Accelerometers are an appealing alternative, but require decoding both the behavioral state \citep{Dot:2016} and the dive type \citep{Hastie:2006} of the animal in order to effectively estimate energy expenditure. The CarHHMM-DFT solves this problem by decoding both behaviors simultaneously at different scales. It also describes the distribution of acceleration within each hidden state. All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.

While we combine the HMM models of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models. For example, HMMs require observations to be taken at regular intervals. Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue, but we do not describe them in detail here. This is because most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit. In addition, we show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions. This implies that a CarHMM is interpretable like a continuous-time model, but as easy to fit as a traditional HMM.

Another consideration when using HMMs to build complex models is the importance of effectively fit the data in question while avoiding \textit{over}-fitting or computationally expensive parameter estimation. One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states. Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states. However, this model is not sufficient if ecologists wish to understand to joint relationship between the dive type and intra-dive behaviour.

@@

Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models. When model complexity is kept under control, a hierarchical model constructed using this processes can be both highly flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale. It is our hope that researchers adapt this methodology to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.
